Location and Scale-Invariant Power Transformations for Transforming Data to Normality

### 2024-08-27 manuscript/manuscript.Rmd
Updates to text.

---
title: "Location and Scale-Invariant Power Transformations for Transforming Data to Normality"
author: "Alex Zwanenburg"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: lualatex
bibliography: refs.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{booktabs}
- \DeclareMathOperator*{\argmax}{argmax}
- \DeclareMathOperator*{\argmin}{argmin}
- \DeclareMathOperator{\sgn}{sgn}
---

@@

Numerical variables in datasets may strongly deviate from normal distributions, e.g. by being skewed. This may complicate further analysis. Power transformations [@Tukey1957-rt] can help improve normality of such features. The two most commonly used transformations are that of @Box1964-mz and @Yeo2000-vw. The Box-Cox transformation of a feature value $x_i$ of feature $\mathbf{X}=\left\{x_1, x_2, \ldots, x_n \right\}$ under the transformation parameter $\lambda$ is defined as:

@@

```{=latex}
\begin{equation}
\label{eqn:yeo-johnson-original}
\phi_{\text{YJ}}^\lambda (x_i) = 
\begin{cases}
\left( \left( 1 + x_i \right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i \geq 0\\
\log(1 + x_i) & \text{if } \lambda = 0 \text{ and } x_i \geq 0\\
-\left( \left( 1 - x_i\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i < 0\\
-\log(1 - x_i) & \text{if } \lambda = 2 \text{ and } x_i < 0
\end{cases}
\end{equation}
```
The $\lambda$-parameter is typically optimised using maximum likelihood estimation under the assumption that the transformed feature is normally distributed. As noted by Raymaekers and Rousseeuw, this approach is very sensitive to the presence of outliers, and robust versions of Box-Cox and Yeo-Johnson transformations were devised [@Raymaekers2021-kq].

Power transformation does not guarantee that transformed variables are normally distributed. In fact, depending on location and scale of the variable, power transformation may decrease normality, as shown in Figure \ref{fig:decreased-normality}. This is problematic for automated power transformations, for three reasons. First, normality is often desired in machine learning applications. Second, a large negative or positive $\lambda$-parameter may lead to numeric issues. Third, we currently do not have a test to automatically reject poor transformations. Statistical tests for normality exist, such as the Shapiro-Wilk test [@Shapiro1965-zd]. However, given sufficiently large sample sizes, these tests can detect trivial deviations from normality and are therefore not well-suited for applications of automated power transformations in practice.

```{r decreased-normality, echo=FALSE, fig.cap=cap, fig.height=4.0, warning=FALSE}
cap <- paste0(
  "Power transformations may reduce normality depending on location. ",
  "Variable $x$ consists of $10000$ values randomly drawn from a normal distribution. ",
  "For the left-hand panels, that distribution is parameterised as $\\mathcal{N}(\\mu, 1)$, ",
  "whereas for the right-hand row, the distribution is $\\mathcal{N}(0, \\sigma)$. " ,
  "For both Box-Cox and Yeo-Johnson transformations, we would expect $\\lambda = 1$, since $x$ is already normally distributed. ",
  "The figure shows that instead, $\\lambda$ is strongly dependent on the location $\\mu$, ",
  "and breaks down due to numerical issues for $\\mu > 5 \\cdot 10^3$. ",
  "Here, scale did not affect Box-Cox and Yeo-Johnson transformations of variables drawn from normal distributions."
)

@@

>>>>

---
title: "Location and Scale-Invariant Power Transformations for Transforming Data to Normality"
author: "Alex Zwanenburg, Steffen Löck"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: lualatex
bibliography: refs.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{booktabs}
- \DeclareMathOperator*{\argmax}{argmax}
- \DeclareMathOperator*{\argmin}{argmin}
- \DeclareMathOperator{\sgn}{sgn}
---

@@

Many statistical and several machine learning methods assume normality of the underlying data. However, numerical features in datasets may strongly deviate from normal distributions, e.g. by being skewed. Power transformations aim to stabilise variance and improve normality of such features [@Bartlett1947-rx, @Tukey1957-rt]. The two most commonly used transformations are that of @Box1964-mz and @Yeo2000-vw. The Box-Cox transformation of a feature value $x_i$ of feature $\mathbf{X}=\left\{x_1, x_2, \ldots, x_n \right\}$ under the transformation parameter $\lambda$ is defined as:

@@

```{=latex}
\begin{equation}
\label{eqn:yeo-johnson-original}
\phi_{\text{YJ}}^\lambda (x_i) = 
\begin{cases}
\left( \left( 1 + x_i \right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i \geq 0\\
\log(1 + x_i) & \text{if } \lambda = 0 \text{ and } x_i \geq 0\\
-\left( \left( 1 - x_i\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i < 0\\
-\log(1 - x_i) & \text{if } \lambda = 2 \text{ and } x_i < 0
\end{cases}
\end{equation}
```
The $\lambda$-parameter is typically optimised using maximum likelihood estimation under the assumption that the transformed feature is normally distributed. As noted by Raymaekers and Rousseeuw, this approach is sensitive to outliers, and robust versions of Box-Cox and Yeo-Johnson transformations were devised [@Raymaekers2021-kq].

Power transformation does not guarantee that transformed variables are normally distributed.
Depending on location and scale of a feature and the presence of outliers, power transformation may decrease normality, as shown in Figure \ref{fig:decreased-normality}.
This is problematic when normality of the transformed feature is not checked, e.g. in automated power transformation in machine learning workflows.
Several issues may arise.
First, some machine learning methods assume normality.
Second, a large negative or positive $\lambda$-parameter may lead to numeric issues.
Third, we currently do not have a test to automatically reject poor transformations to normality.
Statistical tests for normality exist, such as the Shapiro-Wilk test [@Shapiro1965-zd].
However, given sufficiently large sample sizes such tests can detect trivial deviations from normality.

```{r decreased-normality, echo=FALSE, fig.cap=cap, fig.height=2.0, warning=FALSE}
cap <- paste0(
  "Power transformations may reduce normality depending on location. ",
  "Variable $x$ consists of $10000$ values randomly drawn from a normal distribution. ",
  "For the left-hand panels, that distribution is parameterised as $\\mathcal{N}(\\mu, 1)$, ",
  "whereas for the right-hand row, the distribution is $\\mathcal{N}(0, \\sigma)$. " ,
  "For both Box-Cox and Yeo-Johnson transformations, we would expect $\\lambda = 1$, since $x$ is already normally distributed. ",
  "The figure shows that instead, $\\lambda$ is strongly dependent on the location $\\mu$, ",
  "and breaks down due to numerical issues for $\\mu > 5 \\cdot 10^3$. ",
  "Here, scale did not affect Box-Cox and Yeo-Johnson transformations of variables drawn from normal distributions."
)

@@

-   We derive the maximum likelihood criterion for location- and scale-invariant Box-Cox and Yeo-Johnson transformations to allow for optimising transformation parameters.
